 

### 2021-07-05 vertical-thermal-review.Rmd
cut parts of Photosynthesis

~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig. `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie. boundary layer conductance) and latent heat loss (via $gs$ ie. Stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].

@@

Leaf size and physiology [see next section] influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$. Smaller leaves have thinner boundary layer that enables coupling with $T_{air}$ through higher $ga$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig. `r fig_leaf_T`a). Smaller leaves can also heat and cool faster and with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018]. However, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [@leighThickLeavesAvoid2012;@bonan_ecological_2016]. Sun leaves, such as in the upper canopies, tend to have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves. But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021]. However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in $T_{air}$, $T_{leaf}$-$T_{air}$ can become more elevated than sun leaves (Fig. `r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].

**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, and in turn the vertical gradient of $T_{air}$ depends on canopy structure and $LAI$. Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate. The intensity of incoming radiation and $VPD$ can also limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021]. On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013]. 
It is also important to note that, NEON sites (in Fig. `r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures. But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].~~

@@

### In situ observations
**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

The extent of this buffering varies across forest types.
*Available published data suggest that upper canopies of temperate-deciduous species can have greater $T_{leaf} - T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig. `r fig_NEON_vertical`).* 
Because needle-leaves have lower $g_a$ than broad leaves, needle-leaf conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].

@@

In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubisco, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P, and water availability.

Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008]. Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II], leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].

>>>>

~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig. `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie. boundary layer conductance) and latent heat loss (via $gs$ ie. stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].

@@

Leaf size and physiology [see next section] influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$. Smaller leaves have thinner boundary layer and higher $ga$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig. `r fig_leaf_T` a). Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018]. Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [(Fig. `r fig_leaf_T` d; @leighThickLeavesAvoid2012;@bonan_ecological_2016]. Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves. However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021]. However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in radiation or $T_{air}$, $T_{leaf}$-$T_{air}$ can easily become more elevated than sun leaves (Fig. `r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].

**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate. The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$. Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021]. However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).
On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].

NEON sites (in Fig. `r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures. But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].~~

@@

### In situ observations
**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
In closed-canopy forests, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions [Table 1; Table 2]. Light saturation levels differ between leaves and canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P, and water availability.

Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008]. Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions [Tan et al. 2017; @slotSituTemperatureResponse2017]. Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II], leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{Opt}$ across sites and seasons (e.g. Tan et al. 2017; Kumarathunge et al. 2019), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of[ @slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
